---
title: Regional association plots of eQTLs
author: "Silva Kasela"
date: "8/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path = here::here("eqtl", "cis", "fig_regional_plots/"), dev = c('png', 'pdf'))

library(here)
library(arrow)
library(dplyr)
library(ggplot2)
library(Gviz)
library(biomaRt)
library(kableExtra)

# ggplot2 theme
theme_set(theme_classic() +
            theme(plot.title = element_text(size = 14),
                  plot.subtitle = element_text(size = 12),
                  axis.title = element_text(size = 12),
                  axis.title.y = element_text(vjust = 2),
                  axis.title.x = element_text(vjust = -0.75),
                  axis.text = element_text(size = 11),
                  strip.text = element_text(size = 11),
                  legend.title = element_text(size = 12),
                  legend.text = element_text(size = 11)))


```

```{r}
eqtl <- read.table(here("eqtl", "cis", "spiromics.cis_eqtl.egenes.txt.gz"), header = T, sep = "\t", stringsAsFactors = F)

```

### Regional association plots of eQTLs for ACE2 and TMPRSS2

```{r}
gene_df <- data.frame("gene_id" = c("ENSG00000130234", "ENSG00000184012"),
                      "gene_name" = c("ACE2", "TMPRSS2"),
                      stringsAsFactors = F)

```

```{r}
eqtl %>% 
  filter(gene_id %in% gene_df$gene_id) %>% 
  kable(row.names = FALSE) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F, font_size = 14)

```

```{r}
chr <- c("chrX", "chr21")

eqtl_data_list <- lapply(chr, function(chr) {
  file <- here("eqtl", "cis", "allpairs", paste0("spiromics.cis_qtl_pairs.", chr, ".parquet"))
  
  dat <- read_parquet(file) %>% 
    filter(phenotype_id %in% gene_df$gene_id) %>% 
    mutate(gene_name = gene_df[match(phenotype_id, gene_df$gene_id), "gene_name"])
  
  dat$pos <- sapply(dat$variant_id, function(x) unlist(strsplit(x, "_"))[2])
  dat$pos <- as.numeric(dat$pos)
  return(dat)
})

```

```{r fig-ace2-tmprss2, fig.width=5, fig.height=4, fig.show='hold'}
for (i in 1:length(eqtl_data_list)) {
  eqtl_data <- eqtl_data_list[[i]]
  x_points <- seq(round(median(eqtl_data$pos) - 1e6, -6), median(eqtl_data$pos) + 1e6, 1e6/2) # define x-axis breaks

  # eGene significance level
  stats_egene <- eqtl[match(eqtl_data$gene_name[1], eqtl$gene_name), c("gene_name", "pval_nominal_threshold")]
  
  g <- ggplot(eqtl_data, aes(x = pos, y = -log10(pval_nominal))) +
    labs(title = paste0("cis-eQTLs for ", eqtl_data$gene_name[1], " in bronchial epithelium"),
           x = paste0("Position on ", chr[i], " (Mb)"),
           y = bquote(-log[10] ~ "(eQTL P-value)")) +
    geom_point(col = "grey20") +
    #geom_hline(yintercept = -log10(1e-4), col = "indianred", lty = 2) +
    geom_hline(data = stats_egene, aes(yintercept = -log10(pval_nominal_threshold)), col = "indianred", lty = 2) +
    scale_x_continuous(breaks = x_points, labels = x_points/1e6, expand = expansion(mult = c(0.01, 0.01), add = c(0, 0))) +
    scale_y_continuous(limits = c(0, 6), expand = expansion(mult = c(0.05, 0.2), add = c(0, 0)))
  print(g)
}
  
```

* Gene track

```{r fig-ace2-tmprss2-gene-track, fig.width=10, fig.height=3, fig.show='hold'}
for (i in 1:length(eqtl_data_list)) {
  eqtl_data <- eqtl_data_list[[i]]
  # Gene diagrams
  gen <- "hg38"
  xmin <- min(eqtl_data$pos)
  xmax <- max(eqtl_data$pos)
  
  # ideogram
  itrack <- IdeogramTrack(genome = gen, chromosome = chr[i])
  
  # coordinates
  gtrack <- GenomeAxisTrack()
  
  ## highlight lead gwas variant
  #atrack <- AnnotationTrack(genome = gen, chromosome = chr, start = gwas_pos, end = gwas_pos + 1, name = "SNP", col = "royalblue",  cex.title = 1.1)
  
  # ensembl genes track
  bm <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl")
  # limit the returned gene models to RefSeq models only - filter=list(with_refseq_mrna=TRUE), 
  biomtrack <- BiomartGeneRegionTrack(genome = gen, chromosome = chr[i], start = xmin, end = xmax, name = "Genes", biomart = bm, cex.title = 1.1, stacking = "squish")
  
  # plot the union of all the exons as a sort of meta-transcript
  plotTracks(list(itrack, gtrack, biomtrack), collapseTranscripts = "meta", transcriptAnnotation = "symbol", stackHeight = 0.25, from = xmin, to = xmax, cex.main = 1.1)
}

```


### Regional association plots of eQTLs in the chr3 locus associated with COVID-19 GWAS

At locus 3p21.31, the association signal related to COVID-19 spanned the genes SLC6A20, LZTFL1, CCR9, FYCO1, CXCR6, and XCR1

* COVID-19 GWAS data: ANA_B2_V2 - hospitalized covid vs. population (round 3)
    + Lead variant: chr3_45867532_A_G_b38
* COVID-19 GWAS data: ANA_C2_V2 - covid vs. population (round 3)
    + Lead variant: chr3_45867022_C_G_b38
* Ellinghaus et al. 2020, severe COVID-19 with respiratory failure
    + Lead variant: chr3_45834967_G_GA_b38

```{r}
gene_df <- data.frame("gene_id" = c("ENSG00000163817", "ENSG00000163818", "ENSG00000173585", "ENSG00000163820", "ENSG00000172215", "ENSG00000173578"),
                        "gene_name" = c("SLC6A20", "LZTFL1", "CCR9", "FYCO1", "CXCR6", "XCR1"),
                        stringsAsFactors = F)

lead_gwas_id <- c("chr3_45867022_C_G_b38", # ANA_C2_V2
                  "chr3_45867532_A_G_b38", # ANA_B2_V2
                  "chr3_45834967_G_GA_b38") # Ellinghaus et al

```

```{r}
eqtl %>% 
  filter(gene_id %in% gene_df$gene_id) %>% 
  kable(row.names = FALSE) %>% 
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F, font_size = 14)

```

```{r}
chr <- "chr3"
file <- here("eqtl", "cis", "allpairs", paste0("spiromics.cis_qtl_pairs.", chr, ".parquet"))

eqtl_data <- read_parquet(file) %>% 
  filter(phenotype_id %in% gene_df$gene_id) %>% 
  mutate(gene_name = gene_df[match(phenotype_id, gene_df$gene_id), "gene_name"],
         gene_name_f = factor(gene_name, levels = gene_df$gene_name))

eqtl_data$pos <- sapply(eqtl_data$variant_id, function(x) unlist(strsplit(x, "_"))[2])
eqtl_data$pos <- as.numeric(eqtl_data$pos)

```

```{r fig-chr3, fig.width=5, fig.height=12}
x_points <- seq(round(median(eqtl_data$pos) - 1e6, -6), median(eqtl_data$pos) + 1e6, 1e6/2) # define x-axis breaks

# highlight lead GWAS variants
top_variant <- eqtl_data %>%
  filter(variant_id %in% lead_gwas_id)

# eGene significance level
stats_egene <- eqtl[match(gene_df$gene_name, eqtl$gene_name), c("gene_name", "pval_nominal_threshold")]
stats_egene[is.na(stats_egene$pval_nominal_threshold), "pval_nominal_threshold"] <- 1
stats_egene$gene_name_f <- factor(gene_df$gene_name, levels = gene_df$gene_name)

# individual y_max => need to create blank data
min_p <- tapply(eqtl_data$pval_nominal, eqtl_data$gene_name, min)
min_p <- min_p[levels(eqtl_data$gene_name_f)]
blank_data <- data.frame(pos = median(eqtl_data$pos),
                         pval_nominal = c(ifelse(min_p > 10**-6 | is.na(min_p) , 10**-6, min_p), rep(1, length(min_p))),
                         gene_name_f = factor(rep(levels(eqtl_data$gene_name_f), 2), levels = levels(eqtl_data$gene_name_f)),
                         stringsAsFactors = F)
  
ggplot(eqtl_data, aes(x = pos, y = -log10(pval_nominal))) +
  labs(title = "cis-eQTLs in bronchial epithelium",
         x = paste0("Position on ", chr, " (Mb)"),
         y = bquote(-log[10] ~ "(eQTL P-value)")) +
  geom_point(col = "grey20") +
  #geom_hline(yintercept = -log10(1e-4), col = "indianred", lty = 2) +
  geom_hline(data = stats_egene, aes(yintercept = -log10(pval_nominal_threshold)), col = "indianred", lty = 2) +
  geom_point(data = top_variant, aes(col = variant_id)) +
  geom_blank(data = blank_data) +
  scale_x_continuous(breaks = x_points, labels = x_points/1e6, expand = expansion(mult = c(0.01, 0.01), add = c(0, 0))) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.2), add = c(0, 0))) +
  facet_wrap(~gene_name_f, drop = FALSE, nrow = nrow(gene_df), scales = "free_y") +
  theme(legend.position = "top",
        legend.direction = "vertical")

```

* Gene track

```{r fig-chr3-gene-track, fig.width=10, fig.height=3}
# Gene diagrams
gen <- "hg38"
xmin <- min(eqtl_data$pos)
xmax <- max(eqtl_data$pos)

# ideogram
itrack <- IdeogramTrack(genome = gen, chromosome = chr)

# coordinates
gtrack <- GenomeAxisTrack()

## highlight lead gwas variant
#atrack <- AnnotationTrack(genome = gen, chromosome = chr, start = gwas_pos, end = gwas_pos + 1, name = "SNP", col = "royalblue",  cex.title = 1.1)

# ensembl genes track
bm <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl")
# limit the returned gene models to RefSeq models only - filter=list(with_refseq_mrna=TRUE), 
biomtrack <- BiomartGeneRegionTrack(genome = gen, chromosome = chr, start = xmin, end = xmax, name = "Genes", biomart = bm, cex.title = 1.1, stacking = "squish")

# plot the union of all the exons as a sort of meta-transcript
plotTracks(list(itrack, gtrack, biomtrack), collapseTranscripts = "meta", transcriptAnnotation = "symbol", stackHeight = 0.25, from = xmin, to = xmax, cex.main = 1.1)

```

