---
title: "Final list of candidate genes"
author: "Silva Kasela"
date: "5/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path = here::here("candidate_genes", "fig/"), dev = c('png', 'pdf'))
options(knitr.table.format = "html") 

library(here)
library(openxlsx)
library(VennDiagram)
library(dplyr)
library(UpSetR)
library(knitr)
library(kableExtra)

```

```{r read-in-data}
# Candidate genes version 10 - need to filter genes from the Blanco-Melo study for the final list
version <- 10
candidate_genes <- read.table(here("candidate_genes", paste0("covid19_candidate_genes_v", version, ".txt")), header = T, sep = "\t", stringsAsFactors = F)
colnames(candidate_genes)[1] <- "gene_name"

# Gencode annotations
gene_annot <- read.table(here("references", "gencode.v33.annotation.GRCh38.gene_id_name_type.txt"), header = T, sep = "\t", stringsAsFactors = F)

# Genes expressed in spiromics
norm_expr <- data.table::fread(here("expression", "normalized", "spiromics.normalized_expression.bed.gz"), header = T, sep = "\t", stringsAsFactors = F, data.table = F)

```

## Candidate genes studies

Candidate genes from different sources:

* Hoffmann et al. 2020, Cell, _SARS-CoV-2 Cell Entry Depends on ACE2 and TMPRSS2 and Is Blocked by a Clinically Proven Protease Inhibitor_
* Blanco-Melo et al. 2020, Cell, _Imbalanced host response to SARS-CoV-2 drives development of COVID-19_
    + (Title of the pre-print: _SARS-CoV-2 launches a unique transcriptional signature from in vitro, ex vivo, and in vivo systems_)
    + Transcriptional response to SARS-CoV-2 _in vitro_, _ex vivo_, and _in vivo_ models
* Gordon et al. 2020, Nature, _A SARS-CoV-2 protein interaction map reveals targets for drug repurposing_
    + (Title of the pre-print title: _A SARS-CoV-2-Human Protein-Protein Interaction Map Reveals Drug Targets and Potential Drug-Repurposing_)
    + 332 high-confidence SARS-CoV-2-human protein-protein interactions (PPIs)
* Gassen et al. 2020, bioRxiv, _Analysis of SARS-CoV-2-controlled autophagy reveals spermidine, MK-2206, and niclosamide as putative antiviral therapeutics_
* Wang, K. et al. 2020, bioRxiv, _SARS-CoV-2 invades host cells via a novel route: CD147-spike protein_
* COVID-19 Cell Atlas, https://www.covid19cellatlas.org/

```{r}
# Add new titles
for (i in 1:nrow(candidate_genes)) {
  if (grepl("SARS-CoV-2 launches a unique transcriptional signature from in vitro, ex vivo, and in vivo systems", candidate_genes$Source[i])) {
    candidate_genes$Source[i] <- sub("SARS-CoV-2 launches a unique transcriptional signature from in vitro, ex vivo, and in vivo systems",
                                     "Imbalanced host response to SARS-CoV-2 drives development of COVID-19",
                                     candidate_genes$Source[i])
  }
  if (grepl("A SARS-CoV-2-Human Protein-Protein Interaction Map Reveals Drug Targets and Potential Drug-Repurposing", candidate_genes$Source[i])) {
    candidate_genes$Source[i] <- sub("A SARS-CoV-2-Human Protein-Protein Interaction Map Reveals Drug Targets and Potential Drug-Repurposing",
                                     "A SARS-CoV-2 protein interaction map reveals targets for drug repurposing",
                                     candidate_genes$Source[i])
  }
}

refs <- c("Hoffmann et al. 2020" = "SARS-CoV-2 Cell Entry Depends on ACE2 and TMPRSS2 and Is Blocked by a Clinically Proven Protease Inhibitor",
          "Blanco-Melo et al. 2020" = "Imbalanced host response to SARS-CoV-2 drives development of COVID-19",
          "Gordon et al. 2020" = "A SARS-CoV-2 protein interaction map reveals targets for drug repurposing",
          "Gassen et al. 2020" = "Analysis of SARS-CoV-2-controlled autophagy reveals spermidine, MK-2206, and niclosamide as putative antiviral therapeutics",
          "Wang et al. 2020" = "SARS-CoV-2 invades host cells via a novel route: CD147-spike protein",
          "COVID-19 Cell Atlas" = "covid19cellatlas")

candidate_genes$ref <- NULL
for (i in 1:nrow(candidate_genes)) {
  idx <- sapply(refs, function(x) grepl(x, candidate_genes$Source[i]))
  candidate_genes$ref[i] <- paste(names(refs[idx]), collapse = "; ")
}

```

#### Select candidate genes from Blanoc-Melo et al.

Genes from Blanco-Melo et al. 2020  are divided into three groups based on experimental evidence:

* In vitro tissue culture (variety of respiratory cell lines)
    + Human adenocarcinomic alveolar basal epithelial (A549) cells
        + Also, A549 cells supplemented with a vector expressing ACE2 (A549-ACE2), low-MOI infection (MOI, 0.2), because A549 lung alveolar cells are found to be relatively non-permissive to SARS-CoV-2 replication
    + Human adenocarcinomic lung epithelial (Calu-3) cells
* Ex vivo infection of primary cells (normal human bronchial epithelial (NHBE) cells)
    + NHBE cells isolated from a 79-year-old Caucasian female
* In vivo samples derived from both COVID-19 patients (and animals)
    + Compared post-mortem lung samples from two COVID-19-positive patients with biopsied healthy lung tissue from uninfected individuals.

Note about differential expression analysis:

* DE analysis performed by matching each experimental condition with a corresponding mock treated sample that counted for the cell type, collection time, concurrent animal controls, etc.

```{r read-data-blanco-malo}
# Cell line: selecting Calu-3 and A549 ACE2 expressing cells (low MOI)
cellline <- read.xlsx(xlsxFile = here("candidate_genes", "blanco_melo_et_al", "1-s2.0-S009286742030489X-mmc1.xlsx"), sheet = 2, na.strings = "NaN")
# clean data if p-val = 0, then probably a bug (?), replace these with 1
## apply(cellline[, grepl("padj", colnames(cellline))], 2, function(x) sum(x ==0))
## apply(cellline[, grepl("padj", colnames(cellline))], 2, function(x) sum(x ==1))
for (i in 12:21) {
  cellline[,i] <- ifelse(cellline[,i] == 0, 1, cellline[,i])
}

# Primary cells
primary <- read.xlsx(xlsxFile = here("candidate_genes", "blanco_melo_et_al", "1-s2.0-S009286742030489X-mmc2.xlsx"), sheet = 2, na.strings = "NaN")
## apply(primary[, grepl("padj", colnames(primary))], 2, function(x) sum(x ==0))
## apply(primary[, grepl("padj", colnames(primary))], 2, function(x) sum(x ==1))

# Human experiment
human <- read.xlsx(xlsxFile = here("candidate_genes", "blanco_melo_et_al", "1-s2.0-S009286742030489X-mmc4.xlsx"), sheet = 2, na.strings = "NaN")

stopifnot(cellline$GeneName %in% primary$GeneName)
stopifnot(cellline$GeneName %in% human$Gene_name)

```

__1) Overlap between DEGs from different experiments__

* Adjusted p-value < 0.05
    + Adj p-values = 0 in the DE results files from cell lines -> possible bug? -> changing these p-values = 1

_In vitro cell lines kept separately or taking the intersection to form one cluster_

```{r fig-venn-blanco-melo-padj, fig.width=5, fig.height=5, out.width='50%', fig.show='hold'}
cellline_calu3 <- cellline[cellline$`padj_SARS-CoV-2(Calu-3)` < 0.05, ]
cellline_a549_ace2 <- cellline[cellline$`padj_SARS-CoV-2(A549-ACE2)LowMOI` < 0.05, ]
primary <- primary[primary$`padj_SARS-CoV-2` < 0.05, ]
human <- human[!is.na(human$padj) & human$padj < 0.05, ]

x <- list("in vitro\n(Calu-3)" = cellline_calu3$GeneName,
          "in vitro\n(A549-ACE2*)" = cellline_a549_ace2$GeneName,
          "ex vivo" = primary$GeneName,
          "in vivo" = human$Gene_name)
sapply(x, length)

venn_plot <- venn.diagram(
	x = x,
	filename = NULL,
	col = "transparent",
	lty = "dotted",
	lwd = 4,
	fill = c("cornflowerblue", "green", "yellow", "darkorchid1"),
	alpha = 0.50,
	label.col = c("orange", "white", "darkorchid4", "white", "white", "white", "white", "white", "darkblue", "white", "white", "white", "white", "darkgreen", "white"),
	cex = 1.25,
	fontfamily = "serif",
	fontface = "bold",
	cat.col = c("darkblue", "darkgreen", "orange", "darkorchid4"),
	cat.cex = 1.5,
	cat.fontfamily = "serif"
	)
grid.draw(venn_plot)
grid.newpage()

# cell lines together
venn_plot <- venn.diagram(
	x = list(
		"in vitro" = intersect(cellline_calu3$GeneName, cellline_a549_ace2$GeneName),
		"ex vivo" = primary$GeneName,
		"in vivo" = human$Gene_name
		),
	filename = NULL,
	col = "transparent",
	fill = c("red", "blue", "green"),
	alpha = 0.5,
	label.col = c("darkred", "white", "darkblue", "white", "white", "white", "darkgreen"),
	cex = 1.5,
	fontfamily = "serif",
	fontface = "bold",
	cat.default.pos = "text",
	cat.col = c("darkred", "darkblue", "darkgreen"),
	cat.cex = 1.5,
	cat.fontfamily = "serif",
	cat.dist = c(0.06, 0.06, 0.03),
	cat.pos = 0
	)
grid.draw(venn_plot)
grid.newpage()

```

Effects sizes in these datasets:

* Calu-3

```{r}
summary(abs(cellline[cellline$`padj_SARS-CoV-2(Calu-3)` < 0.05, "SARS-CoV-2(Calu-3)_L2FC"]))
```

* A549-ACE2 (LowMOI)

```{r}
summary(abs(cellline[cellline$`padj_SARS-CoV-2(A549-ACE2)LowMOI` < 0.05, "SARS-CoV-2(A549-ACE2)LowMOI_L2FC"]))
```

* Primary cells (NHBE)

```{r}
summary(abs(primary$`SARS-CoV-2_L2FC`))
```

* Human experiment

```{r}
summary(abs(human$log2FoldChange))
```

__2) Overlap between DEGs from different experiments__

* Adjusted p-value < 0.05 & abs(log2 FC) > 0.5 or 1 in primary cell dataset and other datasets, respectively
    + Adj p-values = 0 in the DE results files from cell lines -> possible bug? -> changing these p-values = 1

_In vitro cell lines kept separately or taking the intersection to form one cluster_

```{r fig-venn-blanco-melo-padj-log2fc, fig.width=5, fig.height=5, out.width='50%', fig.show='hold'}
cellline_calu3 <- cellline[cellline$`padj_SARS-CoV-2(Calu-3)` < 0.05 & abs(cellline$`SARS-CoV-2(Calu-3)_L2FC`) > 1, ]
cellline_a549_ace2 <- cellline[cellline$`padj_SARS-CoV-2(A549-ACE2)LowMOI` < 0.05 & abs(cellline$`SARS-CoV-2(A549-ACE2)LowMOI_L2FC`) > 1, ]
primary <- primary[primary$`padj_SARS-CoV-2` < 0.05 & abs(primary$`SARS-CoV-2_L2FC`) > 0.5, ]
human <- human[!is.na(human$padj) & human$padj < 0.05 & abs(human$log2FoldChange) > 1, ]

x <- list("in vitro\n(Calu-3)" = cellline_calu3$GeneName,
          "in vitro\n(A549-ACE2*)" = cellline_a549_ace2$GeneName,
          "ex vivo" = primary$GeneName,
          "in vivo" = human$Gene_name)
sapply(x, length)

venn_plot <- venn.diagram(
	x = x,
	filename = NULL,
	col = "transparent",
	lty = "dotted",
	lwd = 4,
	fill = c("cornflowerblue", "green", "yellow", "darkorchid1"),
	alpha = 0.50,
	label.col = c("orange", "white", "darkorchid4", "white", "white", "white", "white", "white", "darkblue", "white", "white", "white", "white", "darkgreen", "white"),
	cex = 1.25,
	fontfamily = "serif",
	fontface = "bold",
	cat.col = c("darkblue", "darkgreen", "orange", "darkorchid4"),
	cat.cex = 1.5,
	cat.fontfamily = "serif"
	)
grid.draw(venn_plot)
grid.newpage()

# cell lines together
venn_plot <- venn.diagram(
	x = list(
		"in vitro" = intersect(cellline_calu3$GeneName, cellline_a549_ace2$GeneName),
		"ex vivo" = primary$GeneName,
		"in vivo" = human$Gene_name
		),
	filename = NULL,
	col = "transparent",
	fill = c("red", "blue", "green"),
	alpha = 0.5,
	label.col = c("darkred", "white", "darkblue", "white", "white", "white", "darkgreen"),
	cex = 1.5,
	fontfamily = "serif",
	fontface = "bold",
	cat.default.pos = "text",
	cat.col = c("darkred", "darkblue", "darkgreen"),
	cat.cex = 1.5,
	cat.fontfamily = "serif",
	cat.dist = c(0.06, 0.06, 0.03),
	cat.pos = 0
	)
grid.draw(venn_plot)
grid.newpage()

```

#### Selection

For final candidate genes list, select genes that have 

* adjusted P-value < 0.05 in the differential expression analysis from primary cells and either cell lines (Calu-3 or ACE2-expressing A549 cells, low-MOI infection; excluded genes with adjusted P = 0) or samples derived from COVID-19 patients
* log2 fold change > 0.5 in absolute scale in primary cells and log2 fold change > 1 in absolute scale in the other experiment 

```{r}
overlap <- c(intersect(primary$GeneName, cellline_calu3$GeneName),
             intersect(primary$GeneName, cellline_a549_ace2$GeneName),
             intersect(primary$GeneName, human$Gene_name))
overlap <- sort(unique(overlap))
  
overlap_df <- data.frame("gene_name" = overlap,
                        "Source" = "Imbalanced host response to SARS-CoV-2 drives development of COVID-19",
                        "Experimental.Evidence" = NA,
                        "ref" = "Blanco-Melo et al. 2020",
                        stringsAsFactors = F)

overlap_df$Experimental.Evidence <- sapply(overlap_df$gene_name, function(x) {
  if (x %in% cellline_calu3$GeneName & !x %in% cellline_a549_ace2$GeneName & !x %in% human$Gene_name) {
    return("primary_cell;calu3")
  }
  if (!x %in% cellline_calu3$GeneName & x %in% cellline_a549_ace2$GeneName & !x %in% human$Gene_name) {
    return("primary_cell;a549_ace2")
  }
  if (!x %in% cellline_calu3$GeneName & !x %in% cellline_a549_ace2$GeneName & x %in% human$Gene_name) {
    return("primary_cell;covid19_patient")
  }
  if (x %in% cellline_calu3$GeneName & x %in% cellline_a549_ace2$GeneName & !x %in% human$Gene_name) {
    return("primary_cell;calu3;a549_ace2")
  }
  if (x %in% cellline_calu3$GeneName & !x %in% cellline_a549_ace2$GeneName & x %in% human$Gene_name) {
    return("primary_cell;calu3;covid19_patient")
  }
  if (!x %in% cellline_calu3$GeneName & x %in% cellline_a549_ace2$GeneName & x %in% human$Gene_name) {
    return("primary_cell;a549_ace2;covid19_patient")
  }
  if (x %in% cellline_calu3$GeneName & x %in% cellline_a549_ace2$GeneName & x %in% human$Gene_name) {
    return("primary_cell;calu3;a549_ace2;covid19_patient")
  }
})

```

* `r length(overlap)` genes for final candidate gene set

## Overview of the candidate gene list

```{r}
filt <- candidate_genes[!candidate_genes$ref %in% "Blanco-Melo et al. 2020",]
filt$Source <- sub(", Imbalanced host response to SARS-CoV-2 drives development of COVID-19", "", filt$Source)
filt$Source <- sub("Imbalanced host response to SARS-CoV-2 drives development of COVID-19, ", "", filt$Source)
filt$ref <- sub("Blanco-Melo et al. 2020; ", "", filt$ref)
filt$Experimental.Evidence <- ""

# combine two datasets
final <- rbind(filt, overlap_df)

# merge genes that were already present into one
mult <- names(table(final$gene_name)[table(final$gene_name) > 1])
for (g in mult) {
  idx <- which(final$gene_name %in% mult)
  final <- final %>% 
    add_row("gene_name" = g,
            "Source" = paste(final[idx, "Source"], collapse = ", "),
            "Experimental.Evidence" = paste(sapply(final[idx, "Experimental.Evidence"], function(x) {
              if (x %in% "") {
                ""
                } else {
                  x
                }
              }), collapse = ""),
            "ref" = paste(final[idx, "ref"], collapse = ", ")) %>% 
    slice(-idx)
}

rm(mult, filt, candidate_genes)

```

* Note: gene names are matching gencode v33 (changes for genes that are listed by one of their synonym)

```{r rename-candidate-genes}
# gene_id's from gencode
no_match <- final$gene_name[!final$gene_name %in% gene_annot$gene_name]

ensembl <- biomaRt::useMart(biomart = 'ENSEMBL_MART_ENSEMBL', dataset = "hsapiens_gene_ensembl")
#listAttributes(ensembl)
#listFilters(ensembl)
gene_mart <- biomaRt::getBM(attributes = c('ensembl_gene_id', 'hgnc_symbol', 'external_synonym'), 
                            filters = c('external_synonym'), 
                            values = no_match, 
                            mart = ensembl)

# some manual additions (No match for: "LOC100294145" "MGC12916")
gene_mart <- gene_mart %>% 
  add_row("ensembl_gene_id" = "ENSG00000186205", "hgnc_symbol" = "MARC1", "external_synonym" = "MTARC1")

gene_mart$in_genocde <- ifelse(gene_mart$ensembl_gene_id %in% gene_annot$gene_id, 1, 0) 
gene_mart <- gene_mart[gene_mart$in_genocde %in% 1,]
stopifnot(gene_mart$hgnc_symbol %in% gene_annot$gene_name)
stopifnot(gene_mart$ensembl_gene_id == gene_annot[match(gene_mart$hgnc_symbol, gene_annot$gene_name), "gene_id"])

# remove genes that don't have a match in gencode
cat("Genes that do not have a match from gencode - ", paste(no_match[!no_match %in% gene_mart$external_synonym], collapse = ", "), fill = T)
final <- final[!final$gene_name %in% no_match[!no_match %in% gene_mart$external_synonym], ]

# change gene_names
final[match(gene_mart$external_synonym, final$gene_name), "gene_name"] <- gene_mart$hgnc_symbol

```

```{r candidate-gene-names}
stopifnot(final$gene_name %in% gene_annot$gene_name)

final$gene_id <- gene_annot[match(final$gene_name, gene_annot$gene_name), "gene_id"]
final$gene_type <- gene_annot[match(final$gene_name, gene_annot$gene_name), "gene_type"]

# re-order columns
final <- final[, c("gene_id", "gene_name", "gene_type", "ref", "Source", "Experimental.Evidence")]
final <- final[!duplicated(final), ]
colnames(final)[5:6] <- c("ref_title", "experimental_evidence")

final <- final[order(final$gene_name),]

# are these genes expressed in spiromics
final$expressed_in_spiromics <- ifelse(final$gene_id %in% norm_expr$gene_id, 1, 0)

# save
write.table(final, file = here("candidate_genes", paste0("covid19_candidate_genes_v", version + 1, ".txt")), col.names = T, row.names = F, sep = "\t", quote = F)

```

#### Final list of candidate genes

* __`r nrow(final)`__ candidate genes in the filtered list

```{r fig-candidate-genes, fig.width=6, fig.height=4}
listInput <- lapply(refs, function(x) which(grepl(x, final$ref_title)))
upset(fromList(listInput), order.by = "freq", mainbar.y.max = 350,
      sets = c("Gordon et al. 2020", "Blanco-Melo et al. 2020", "COVID-19 Cell Atlas", "Gassen et al. 2020", "Hoffmann et al. 2020", "Wang et al. 2020"),
      keep.order = TRUE,
      mainbar.y.label = "Intersection size of\nCOVID-19 related genes", sets.x.label = "# COVID-19 related genes")

# sapply(listInput, length)

```

Gene biotype

```{r}
addmargins(table("gene biotype" = final$gene_type, "is expressed in spiromics" = final$expressed_in_spiromics))
```

Candidate genes not expressed in SPIROMICS bronchial epithelium samples:

```{r}
kable(final[final$expressed_in_spiromics %in% 0, ], row.names = F) %>% 
   kable_styling(bootstrap_options = c("striped", "condensed"), full_width = F, font_size = 14)
  
```

```{r fig-candidate-genes-only-expressed, fig.width=6, fig.height=4}
final <- final[final$expressed_in_spiromics %in% 1, ]
listInput <- lapply(refs, function(x) which(grepl(x, final$ref_title)))
upset(fromList(listInput), order.by = "freq", mainbar.y.max = 350,
      sets = c("Gordon et al. 2020", "Blanco-Melo et al. 2020", "COVID-19 Cell Atlas", "Gassen et al. 2020", "Hoffmann et al. 2020", "Wang et al. 2020"),
      keep.order = TRUE,
      mainbar.y.label = "Intersection size of\nCOVID-19 related genes", sets.x.label = "# COVID-19 related genes")

# sapply(listInput, length)

```
